---
title: "Analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r, echo=FALSE}
library(readr) #for loading Excel files & txt files
library(here) #to set paths
library(dplyr) #for data processing
library(rstatix)
library(ggplot2)
library(corrplot)

```


```{r}
data_spot <- here::here("data","processed_data","cafinal.rds") 
ca <- read_rds(data_spot)
summary(ca)
```

# Walkthrough of the data:

* Outcome: Covid Hospitalization Ratio = covidhosp = # of hospitalized_covid_patients/CumulativeCase) *1000
* Main Predictor: AQI = avi.aqi

Note: All variables are by county

### Demographic Variables:

* TotalPop
* male
* female
* under_18 
* age18to34         
* age35to64           
* age65          
* not_hispanic
* white             
* black              
* AIANA           
* asian           
* pacific_island         
* other            
* mult_race           
* hisp

Note: variables with r in front ar the rate form. Ex. rblack i=  # of black residents in county / county population

Now I am going through each variable 

# Covid Hospitalization Ratio

Filter out counties with missing data
ca1 <- ca %>%
  filter(!County %in% c("Alpine", "Sierra", "Modoc", 	"Sutter", "Lassen", "Yuba"))

As a reminder, this is the number of covid hospitalization / covid cases per 1,000
```{r}
# here is the distribution of the raw numbers
hist(ca$hospitalized_covid_patients)
```

obviously there is a county with an outlier...it appears to be sorted out for the rate version

```{r}
# hist of the ratio
hist(ca$covidhosp)
hist(sqrt(ca$covidhosp))
```

```{r}
# more histograms
# hist(log(ca$covidhosp))
# boxplot(ca$covidhosp, col=terrain.colors(4) )

# log transforming
ca$logcovidhosp<- log(ca$covidhosp)
```

### outlier identification
```{r}
# identify outliers
ca %>% identify_outliers(covidhosp)
#### Sutter, Placer & Yuba....
```
### now checking normality
```{r}
ca %>% shapiro_test(covidhosp)
```

hmmmm pvalue is <0.05 so we CANNOT assume normality... I am going to go for a one way ANOVA test instead of t test.


### one way anova
```{r}
res.aov <- aov(covidhosp ~ County, data = ca)
# Summary of the analysis
summary.aov(res.aov)

```

I am not getting any pvalues...


# Average Aqi
```{r}
# hist of the ratio
hist(ca$av.aqi)
boxplot(ca$av.aqi, col=terrain.colors(4) )
# we hav a couple extremes

# identify outliers
ca %>% identify_outliers(av.aqi)
#### Sutter, Placer & Yuba....

# now checking normality
ca %>% shapiro_test(av.aqi)
# hmmmm pvalue is <0.05 so we CANNOT assume normality... I am going to go for a one way ANOVA test instead of t test.

# one way anova
res.aov <- aov(av.aqi ~ County, data = ca)
# Summary of the analysis
summary.aov(res.aov)

```

### looking at the relationship between the two
```{r}

ggplot(ca, aes(x = covidhosp, y = av.aqi)) + 
  geom_point(color='#2980B9', size = 4) + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE, color='#2C3E50')
```

# Demographic variables
## Sex
```{r}
# hist of the male
hist(ca$rmale)
# we have a couple extremes

ratemat <- ca %>% select(covidhosp, rmale, rfemale) 
ratemat.cor <- cor(ratemat,  method = "pearson", use = "complete.obs")
ratemat.cor
corrplot.mixed(ratemat.cor, order = 'AOE')
```

## Race / Ethnicity
```{r}
# we have a couple extremes
ratemat <- ca1 %>% select(covidhosp, rnot_hispanic, rwhite, rblack, rAIANA, rasian, rpacific_island, rother, rmult_race, rhisp)
ratemat.cor <- cor(ratemat,  method = "pearson", use = "complete.obs")
ratemat.cor
corrplot.mixed(ratemat.cor, order = 'AOE')
# interesting to look at covid hospitalization...
# strongest appears to be with black and AIANA
```

## Age
```{r}
# we have a couple extremes
ratemat <- ca %>% select(covidhosp, runder_18, rage18to34, rage35to64, rage65,)
ratemat.cor <- cor(ratemat,  method = "spearman", use = "complete.obs")
ratemat.cor
corrplot.mixed(ratemat.cor, order = 'AOE')
# not what i expected...
```


# fianl correlation plot
```{r}
ratemat <- ca1 %>% select(av.aqi:rhisp,covidhosp) #rCumulativeCase,rCumulativeDeath, rTotalPop, male, female, under_18, age18to34, age35to64, age65, not_hispanic, white, black, AIANA, asian, pacific_island, other, mult_race, hisp, av.aqi)
ratemat.cor <- Hmisc::rcorr(na.omit(as.matrix(ratemat)))
corrplot(ratemat.cor$r, type="upper", order="hclust", 
         p.mat = ratemat.cor$P, sig.level = 0.05, insig = "blank")
ratemat.cor$pval

mcor<-round(cor(ratemat, method = "pearson", use = "complete.obs"),2)

mcor
as.data.frame( as.table(mcor) )
pearsonlist <- mcor$covidhosp
tbl <- as.dt(mcor)
```


 
 
# linear modeling

## Variables interested in modeling

Main variable: AQI

Other demographic variables:

- 65+ years
- <18 years
- Female
- White 
- Black
- Asian
- Hispanic

library(reshape2)

## Running some individual lm models 

```{r}
ca2 <- ca1 %>% select(covidhosp, rhisp, rwhite, rblack, rasian, rfemale, runder_18, rage65, av.aqi)
ca2 = melt(ca2, id.vars='covidhosp')
ggplot(ca2) +
  geom_jitter(aes(value,covidhosp, colour=variable),) + geom_smooth(aes(value,covidhosp, colour=variable), method=lm, se=FALSE) +
  facet_wrap(~variable, scales="free_x") +
  labs(x = "Covid Hispitalization Rate per 1,000", y = "Vairable Rate") + 
  theme_minimal() 


```

Should we exclude % white?


### Running models

```{r}
lm1 = lm(covidhosp ~ av.aqi, data = ca1)
summary(lm1)

lm2 = lm(covidhosp ~ av.aqi + rhisp + rwhite + rblack + rasian + rfemale + runder_18 + rage65, data = ca1)
summary(lm2)

lm3 = lm(covidhosp ~ av.aqi + rhisp +  rblack + rasian + runder_18 + rage65, data = ca1)
summary(lm3)

lm4 = lm(covidhosp ~ av.aqi + rhisp +  rblack + runder_18 + rage65, data = ca1)
summary(lm4)

lm5 = lm(covidhosp ~ av.aqi + rhisp +   runder_18 + rage65, data = ca1)

summary(lm5)

# evaluating model 3
qqnorm(lm3$residuals)
qqline(lm3$residuals)

ggplot(data = lm3, aes(x = .fitted, y = .resid))+
    geom_jitter()+
    geom_hline(yintercept = 0, linetype = "dashed")+
    xlab("Fitted Values")+
    ylab("Residuals")
# hmmmmmmm not sure if I like this
hist(lm3$residuals)


```































